---
title: "R Notebook"
output: html_notebook
---

1. Investigate to begin with the effect of prednison by considering estimates of the survival functions for the treatment groups and by performing a log-rank test.

```{r}
source("andersen-plot.R")
library("survival")

S = Surv(cirrhosis[,3],event = cirrhosis[,2],type = "right")
survfit(S~V4,cirrhosis)
plot(survfit(S~V4,cirrhosis),lty=c(1,2))
survdiff(S ~ V4,cirrhosis)
```

2. Fit a Cox proportional model with all covariates (main effects) included.
Perform a careful assessment of the fitted model (including “Andersen”
plots). Also check whether there are any problematic issues with the
covariates (use histograms, boxplots,...).
```{r}
cox.fit = coxph(S~strata(V4)+V5+V6+V7+V8,cirrhosis)
cox.surv = survfit(cox.fit)
andersen.plot(cox.surv)
summary(cox.fit)
coxsnell = survfit(Surv(cirrhosis$V4-residuals(cox.fit),cirrhosis$V4)~1)
plot(log(coxsnell$time),log(-log(coxsnell$surv)))
res.dev = residuals(cox.fit,type="deviance")
{
  boxplot(res.dev~cirrhosis[,4])
  boxplot(res.dev~cirrhosis[,5])
  boxplot(res.dev~cirrhosis[,6])
  plot(cirrhosis[,7],res.dev)
  plot(cirrhosis[,8],res.dev)
  
}

dfbetas = residuals(cox.fit,type = "dfbetas")
for(i in 1:4){
  plot(res.dev,dfbetas[,i])
}

plot(cox.zph(cox.fit))

```
3. Within the framework of the fitted Cox regression model, test the null
hypothesis of no effect of prednison treatment.
```{r}
andersen.plot(cox.surv)
coxph(S~V4+V5+V6+V7+V8,cirrhosis)
```


4. Investigate whether some kind of transformations should be applied to
the continuous covariates age and prothrombin. Here you can use martingale residuals, discretization of the covariates, comparison of models obtained with different transformations etc.
```{r}
cox.pro = coxph(S~V5+V6+V7,cirrhosis)
summary(lm(residuals(cox.pro)~cirrhosis$V8))
d = cirrhosis$V8^2
summary(lm(residuals(cox.pro)~d))
plot(cirrhosis$V8,residuals(cox.pro))

cox.age = coxph(S~V5+V6+V8,cirrhosis)
summary(lm(residuals(cox.age)~cirrhosis$V7))
d = cirrhosis$V7^2
summary(lm(residuals(cox.age)~d))
plot(cirrhosis$V7,residuals(cox.age))
```
5. Investigate whether there are indications of effects of interaction between treatment and the other covariates (this is a kind of data mining/exploratory analysis).
```{r}
summary(lm(V6~V4,cirrhosis))
summary(lm(V7~V4,cirrhosis))
summary(lm(V8~V4,cirrhosis))
coxph(S~V4+V5+V7,cirrhosis)
```
6. Give an estimate for the survival function for a prednison treated man,
who at the start of the treatment was 57, had Ascites=little, and a
prothrombin index equal to 85. What is the probability that he will
survive to be 65 ?
```{r}
cox.pred = coxph(S~V5+V6+V7+V8,cirrhosis,subset=(V4==1))
cox.pred$coefficients[1]
plot(survfit(cox.pred)*cox.pred$coefficients[1])
survfit(cox.pred)
```


